Investigating the outer reaches of the Sculptor and Ursa Minor dwarf galaxies

### 2025-07-15 thesis/methods.tex
major thesis restructure

In this section, we discuss our simulation methods from orbital
estimation and initial conditions to post-processing and software.

We use Agama \citep{agama} for potential specification and initial
conditions, and Gadget-4 \citep{gadget4} for orbit integration and dark
matter simulations (with a custom patch to integrate with Agama).

\section{Milky Way Potential}\label{milky-way-potential}

fig.~\ref{fig:v_circ_potential} plots the circular velocity profiles of
each component of our fiducially potential. We adopt the potential
described in \citet{EP2020}, an analytic approximation of
\citet{mcmillan2011}. The potential consisting of a stellar Bulge, a
thin and thick disk, and a dark matter NFW halo. Especially in the
regime where Sculptor and Ursa Minor orbit, the dark matter halo is the
dominant component of the potential. The details of the stellar disk and
bulge are less important.

The galactic bulge is described by a \citet{hernquist1990} potential.
The potential is

\begin{equation}
\Phi(r) = - \frac{GM}{r + a}
\end{equation} where \(a=1.3\,{\rm kpc}\) is the scale radius and
\(M=2.1 \times 10^{10}\,\Mo\) is the total mass.

The thin and thick disks are represented with the
\citet{miyamoto+nagai1975} cylindrical potential: \begin{equation}
\Phi(R, z) = \frac{-GM}{\left(R^2 + \left[a + \sqrt{z^2 + b^2}\right]^{2}\right)^{1/2}}
\end{equation}

where \(a\) is the disc radial scale length, \(b\) is the scale height,
and \(M\) is the total mass of the disk. For the thin disk,
\(a=3.944\,\)kpc, \(b=0.311\,\)kpc,
\(M=3.944\times10^{10}\,\)M\(_\odot\). For the thick disk,
\(a=4.4\,\)kpc, \(b=0.92\,\)kpc, and \(M=2\times10^{10}\,\)M\(_\odot\).

The halo is a NFW dark matter halo (REF) with
\(M_s=79.5\times10^{10}\,\)M\(_\odot\). and \(r_s = 20.2\,\)kpc.

Variations to the potential of the inner disk (exclusion of a bar)
should minimally affect our results as no orbit we consider reaches less
than \textasciitilde15 kpc of the MW centre. We exclude the mass
evolution of the halo from this analysis. Over \(10\,\)Gyr, this would
be fairly significant (factor of \(\sim 2\)in MW mass, REF) but since we
want to determine the upper limit of tidal effects, it is safe to
neglect this.

Finally, in chapter REF, we consider the influence of the large Milky
Cloud on the orbits and evolution of Scl. We adopt the
\citet{vasiliev2024} multipole approximation of an N-body simulation of
the LMC and MW. Their initial conditions are

\begin{itemize}
\tightlist
\item
  MW halo:
\item
  MW bulge (static):
\item
  MW disk (static):
\item
  LMC halo:
\end{itemize}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/v_circ_potential.pdf}}
\caption[Circular velocity of potential]{Circular velocity profile of
\citet{EP2020} potential. We also show the MW and LMC potential from
\citet{vasiliev2024} for their L3M11 model.}\label{fig:v_circ_potential}
\end{figure}

\section{Orbital Estimation}\label{orbital-estimation}

To estimate the orbits of a dwarf galaxy, we perform a Monte Carlo
sampling of the present-day observables. We integrate each sampled
observable back in time for 10 Gyr in Gadget as massless point particles
outputting positions every 5Myr (with otherwise similar parameters to
n-body runs below.)

To convert from Gaia to galactocentric coordinates, we use the
\citet{astropy} v4 galactocentric frame. This frame assumes the galactic
centre (Sagittarius A*) appears:

\begin{itemize}
\tightlist
\item
  J2000 `ra' = 17h45m37.224s, `dec' = -28°56'10.23'',
  \(\mu_{\alpha*}=-3.151\pm0.018\) mas/yr, \(\mu_\delta=-5.547\pm0.026\)
  from the appendix and Table 2 of \citet{reid+brunthaler2004}.
\item
  distance = \(8.122\pm0.033\,\)kpc, solar radial velocity =
  \(11 + 1.9 \pm 3\) km/s. from \citet{gravitycollaboration+2018}.
\end{itemize}

Finally, adding that the sun is \(20.8\pm0.3\,\)pc above the disk from
bennett+bovy2019, and using the procedure described in
\citet{drimmel+poggio2018}, the solar velocity relative to the galactic
rest frame is

\begin{itemize}
\tightlist
\item
  `v\_sun' = \([-12.9 \pm 3.0, 245.6 \pm 1.4, 7.78 \pm 0.08]\) km/s
\end{itemize}

We select the initial conditions from the location of the first
apocentre of a selected orbit.

While we assume that the galaxy can be described as a point particle,
not subject to (internal) dynamical friction, this approximation works
well if the galaxy is not strongly affected by tides (like Scl).
Analytic approximations of dynamical friction tend to be inadequate to
more accurately model the N-body trajectory.

\section{N-Body Modelling}\label{n-body-modelling}

@@

We use Agama \citep{agama} to generate initial conditions. We initially
assume galaxies are described by an NFW dark matter potential (REF) and
the stars are merely collisionless tracers embedded in this potential
(added on in post-processing). The density is exponentially truncated
with a profile
\begin{equation}\protect\phantomsection\label{eq:trunc_nfw}{
\rho_{\rm tNFW} = e^{-r/r_t}\ \rho_{\rm NFW}(t)
}\end{equation} where we adopt \(r_t = 100 r_s\) or approximately 10
times \(r_{200}\).

@@

To ensure that the initial conditionss of the simulation are dynamically
relaxed and well-converged, we run the simulation in isolation (no
external potential) for 5 Gyr (or approximately the crossing timescale).
Our fiducial isolation halo uses \(r_s=2.76\) kpc and
\(M_s = 0.29 \times 10^{10}\) Msun, but can be easily rescaled for any
length or mass scale.

We adopt a softening length of \begin{equation}
h_{\rm grav} = 0.014 \left(\frac{r_s}{2.76\,{\rm kpc}}\right)\left(\frac{N}{10^7}\right)^{-1/2}.
\end{equation} See appendix REF for a discussion of this choice, which
is similar to the \citet{power+2003} suggested softening.

We use the relative tree opening criterion with the accuracy parameter
set to 0.005, and adaptive time stepping with integration accuracy set
to 0.01.

@@

As discussed in \citet{power+2003}, the region where density is
converged is related to the region which becomes collisionally relaxed
over the time of the universe (so about 5-10 Gyr). (i.e.~where the
collisionless assumption breaks down). The relaxation timescale is given
by \(t_{\rm relax} = t_{\rm circ} N(r) / 8\ln N(r)\) or otherwise,
\begin{equation}
t_{\rm relax}(r) := t_{\rm circ}(r) \frac{N(r)}{8\,\ln N(r)}
= {t_{\rm circ}(r_{200})} \frac{\sqrt{200}}{8} \frac{N(r)}{\ln N(r)} \left(\frac{\bar \rho (r)}{\rho_{\textrm crit}}\right)^{-1/2}
\end{equation} This works out to be about 6-10 times (increasing with
particle number) our adopted softening length for NFW halos given our
assumptions. As such, at full resolution, we can only trust density
profiles down to \(\sim10\epsilon\), just enough to resolve stellar
density profiles.

Note that by decreasing the truncation radius, the effective particle
number is improved, so future experiments could be less generous with
this parameter to improve computational performance (order 30\% better
resolution).

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/iso_converg_num.pdf}}
\caption[Numerical halo convergence]{Numerical convergence test for
circular velocity as a function of log radius for simulations with
different total numbers of particles in isolation. Residuals in lower
panel are relative to NFW. The initial conditions are dotted and the
converged radius is marked by arrows
(REF).}\label{fig:numerical_convergance}
\end{figure}

@@

To perform the simulations of a given galaxy in a given potential, we
centre the isolation run's final snapshot and place the dwarf galaxy in
the specified orbit in the given potential. We typically run the
simulation for 10 Gyr, which allows us to orbit slightly past the
expected initial conditions.

\section{Post Processing}\label{post-processing}

\subsection{Centring}\label{centring}

@@

\begin{itemize}
\tightlist
\item
  Circular velocity is computed assuming spherical symmetry and only
  shown for every 200th particle (ranked from the centre outwards)
\end{itemize}

\begin{equation}
v_{\rm circ}(r) = \sqrt{\frac{GM(r)}{r}}
\end{equation}

\begin{itemize}
\tightlist
\item
  \(v_{\rm circ\ max}\) and \(r_{\rm circ,\ max}\) is found by
  least-squares fitting the NFW functional velocity form to the points
  of the velocity profile that have the 10\% highest velocities. This is
  not a necessarily good fit, especially as the halo becomes stripped,
  but accurate enough to find a reliable maximum.
\end{itemize}

\subsection{Stellar Probabilities}\label{stellar-probabilities}

We assign stellar probabilities (assuming spherical, isotropic symmetry)
via Eddington inversion.

Let \(\Psi\) be the potential (normalized to vanish at infinity) and
\({\cal E}\) is the binding energy \({\cal E} = \Psi - 1/2 v^2\). If we
know \(f({\cal E})\), the distribution function (phase-space density in
energy), then we assign the stellar weight for a given particle with
energy \({\cal E}\) is \begin{equation}
P_\star({\cal E}) = \frac{f_\star({\cal E})}{f_{\rm halo}({\cal E})}.
\end{equation} While \(f({\cal E})\) is a phase-space density, the
differential energy distribution includes an additional \(g({\cal E})\)
occupation term (BTXXX).

We use Eddington inversion to find the distribution function, (eq.
4-140b in BT87) \begin{equation}
f({\cal E}) = \frac{1}{\sqrt{8}\, \pi^2}\left( \int_0^{\cal E} \frac{d^2\rho}{d\Psi^2} \frac{1}{\sqrt{{\cal E} - \Psi}}\ d\Psi + \frac{1}{\sqrt{\cal E}} \left(\frac{d\rho}{d\Psi}\right)_{\Psi=0} \right).
\end{equation}

In practice the right, boundary term is zero as \(\Psi \to 0\) as
\(r\to\infty\), and if \(\rho \propto r^{-n}\) at large \(r\) and
\(\Psi \sim r^{-1}\) then \(d\rho / d\Psi \sim r^{-n+1}\) which goes to
zero provided that \(n > 1\).

\(\Psi\) is taken as known from the underlying assumed analytic dark
matter potential. \(\rho_\star\) is pre-specified or calculated via Abel
integration / deprojection

@@

The 2D exponential stellar profile is given by \begin{equation}
\Sigma_\star(R) = A e^{-R / R_s}
\end{equation} where \(A\) is some normalization and \(R_s\) is the
exponential scale radius. The 3D deprojected profile is found through
abel inversion (e.g.~Rapha\ldots), i.e. \begin{equation}
\rho_\star (r) =- \frac{1}{\pi}\int_r^\infty \frac{d\Sigma}{dR} \frac{1}{\sqrt{R^2 - r^2}} dR  = \frac{\Sigma_0}{\pi R_s^2}\,K_0(r/R_s)
\end{equation} where \(K_0\) is the 0th order modified Bessel function
of the second type.

\subsubsection{Plummer}\label{plummer}

A Plummer profile is defined by \begin{equation}
\Sigma(R) = \frac{M}{4\pi R_s^2} \left(1 + \frac{R^2}{R_s^2}\right)^{-2} ,
\end{equation} where \(M\) is the total mass and \(R_s\) is the
characteristic scale radius. The density is \begin{equation}
\rho(r) = \frac{3M}{4\pi\,R_s^3} \left(1 + \frac{r^2}{R_s^2}\right)^{-5/2}.
\end{equation}

@@

\begin{itemize}
\tightlist
\item
  Stellar velocity dispersion is calculated for all stars within 1kpc
  (3D) of the centre by assuming
  \(\sigma_{\rm los} = \sqrt{\sigma_x^2 + \sigma_y^2 + \sigma_z^2}/ \sqrt 3\),
  i.e.~isotropic velocity dispersion
\item
  We calculate density profiles similar to stars and assume a constant
  bin width in \(\log R\) of \(2 {\rm IQR} / \sqrt[3]{N}\)
  (Freedman-Diaconis prescription). \(R\) may be either the cylendrical
  radius in \(x-y\) or the on-sky project \(\xi, \eta\) tangent plane
  coordinates.
\item
  Because our dwarfs are assumed to be spherical/isotropic, we retain
  this assumption when calculating predicted density profiles.
\end{itemize}

>>>>

In this section, we discuss our methods used for the following two
chapters. We utilize a typical analytic, idealized Milky Way potential
based on \citet{mcmillan2011}. Our simulations are collisionless cold
dark matter N-body simulations integrated with Gadget-4. We check for
numerical convergence and stability of initial conditions in isolation.

\section{Milky Way potential}\label{milky-way-potential}

We adopt the Milky Way potential described in \citet{EP2020}, an
analytic approximation of \citet{mcmillan2011}.
Fig.~\ref{fig:v_circ_potential} plots the circular velocity profiles of
each component of our fiducially potential. The potential consisting of
a stellar Bulge, a thin and thick disk, and a dark matter NFW halo.
Especially in the regime where Sculptor and Ursa Minor orbit, the dark
matter halo is the dominant component of the potential. The details of
the stellar disk and bulge are less important.

The potential is specified as follows. The galactic bulge is described
by a \citet{hernquist1990} potential. The potential is

\begin{equation}\protect\phantomsection\label{eq:hernquist}{
\Phi(r) = - \frac{GM}{r + a}
}\end{equation} where \(a=1.3\,{\rm kpc}\) is the scale radius and
\(M=2.1 \times 10^{10}\,\Mo\) is the total mass. The thin and thick
disks are represented with the \citet{miyamoto+nagai1975} cylindrical
potential:

\begin{equation}\protect\phantomsection\label{eq:mn75}{
\Phi(R, z) = \frac{-GM}{\left(R^2 + \left[a + \sqrt{z^2 + b^2}\right]^{2}\right)^{1/2}}
}\end{equation}

where \(a\) is the disc radial scale length, \(b\) is the scale height,
and \(M\) is the total mass of the disk. For the thin disk,
\(a=3.944\,\)kpc, \(b=0.311\,\)kpc,
\(M=3.944\times10^{10}\,\)M\(_\odot\). For the thick disk,
\(a=4.4\,\)kpc, \(b=0.92\,\)kpc, and \(M=2\times10^{10}\,\)M\(_\odot\).
The halo is a NFW dark matter halo (Eq.~\ref{eq:nfw}) with
\(M_s=79.5\times10^{10}\,\)M\(_\odot\). and \(r_s = 20.2\,\)kpc.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/v_circ_potential.png}}
\caption[Circular velocity of potential]{Circular velocity profile of
\citet{EP2020} potential.}\label{fig:v_circ_potential}
\end{figure}

\section{Orbital estimation}\label{orbital-estimation}

To estimate the possible orbits of a dwarf galaxy, we perform a Monte
Carlo sampling of the present-day observables. The present-day position,
distance modulus, LOS velocity, and proper motions are each sampled from
normal distributions given the reported uncertainties in
Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}. We integrate
each sampled observable back in time for 10 Gyr in Gadget as massless
point particles outputting positions every 5Myr (with otherwise similar
parameters to n-body runs below.) When selecting the initial position
and velocity of an N-body model, we select the position and velocity of
the first apocentre occurring after 10 Gyr ago.

We assume the galaxy is represented as a point particle for this
analysis. In detail, (self) dynamical friction likely influences the
orbit, but a point particle is an adequate description.

To convert from Gaia to galactocentric coordinates, we use the Astropy
v4 galactocentric frame \citep{astropycollaboration+2022}. This frame
assumes the galactic centre appears at position
\(\alpha = {\rm 17h45m37.224s}\), \(\delta = -28°56'10.23''\) with
proper motions \(\mu_{\alpha*}=-3.151\pm0.018\ \masyr\) ,
\(\mu_\delta=-5.547\pm0.026 \masyr\) (from the appendix and Table 2 of
\citet{reid+brunthaler2004}). The galactic centre distance is
\(8.122\pm0.033\,\)kpc with a radial velocity = \(11 + 1.9 \pm 3\) km/s.
from \citet{gravitycollaboration+2018}. Finally, adding that the sun is
\(20.8\pm0.3\,\)pc above the disk from \citet{bennett+bovy2019}, and
using the procedure outlined in \citet{drimmel+poggio2018}, the solar
velocity relative to the galactic rest frame is `v\_sun' =
\([-12.9 \pm 3.0, 245.6 \pm 1.4, 7.78 \pm 0.08]\) km/s. The
uncertainties in the reference frame are typically smaller than the
uncertainties on the dwarf galaxy's position and velocity.

\section{N-Body modelling}\label{n-body-modelling}

@@

We use Agama \citep{agama} to generate initial conditions. We initially
assume galaxies are described by an NFW dark matter potential
Eq.~\ref{eq:nfw} and the stars are merely collisionless tracers embedded
in this potential (added on in post-processing). The density is a
cubic-exponentially truncated with a profile
\begin{equation}\protect\phantomsection\label{eq:trunc_nfw}{
\rho_{\rm tNFW} = e^{-(r/r_t)^3}\ \rho_{\rm NFW}(t)
}\end{equation} where we adopt \(r_t = 20 r_s\) or approximately
\(r_{200}\) for our Sculptor-like fiducial halo. Using \(r_{200}\) for
\(r_t\) would depend on the chosen scale of the halo. So our adopted
\(r_t\) is an approximate upper limit of \(r_{200}\) for typical dwarf
galaxy halos \citep{ludlow+2016}. It is unlikely that the outer density
profile of loosely bound particles past \(20\) kpc affects the tidal
evolution of a subhalo.

@@

To ensure that the initial conditionss of the simulation are dynamically
relaxed and well-converged, we run the simulation in isolation (no
external potential) for 5 Gyr (or about 3 times the crossing timescale
at the virial radius). Our fiducial isolation halo uses \(r_s=2.76\) kpc
and \(M_s = 0.29 \times 10^{10}\) Msun, but can be easily rescaled for
any length or mass scale.

For our simulation parameters, we adopt a softening length of
\begin{equation}{
h_{\rm grav} = 0.014 \left(\frac{r_s}{2.76\,{\rm kpc}}\right)\left(\frac{N}{10^7}\right)^{-1/2}.
}\end{equation} See appendix Section~\ref{sec:extra_convergence} for a
discussion of this choice, which is similar to the \citet{power+2003}
suggested softening. We use the relative tree opening criterion with the
accuracy parameter set to 0.005, and adaptive time stepping with
integration accuracy set to 0.01.

@@

As discussed in \citet{power+2003}, the region where density is
converged is related to the region which becomes collisionally relaxed
over the time of the universe (so about 5-10 Gyr). (i.e.~where the
collisionless assumption breaks down). The relaxation timescale is given
by \begin{equation}\protect\phantomsection\label{eq:t_relax}{
t_{\rm relax}(r) := t_{\rm circ}(r) \frac{N(r)}{8\,\ln N(r)}
= {t_{\rm circ}(r_{200})} \frac{\sqrt{200}}{8} \frac{N(r)}{\ln N(r)} \left(\frac{\bar \rho (r)}{\rho_{\textrm crit}}\right)^{-1/2}
}\end{equation} where \(N(r)\) is the number of particles within \(r\),
\(t_{\rm circ}\) is the circular velocity timescale, and \(\bar \rho\)
is the mean interior density \citep{power+2003}. The converged radius
\(r_{\rm relax}(t)\) is the radius above which \(t_{\rm relax} > t\).
Typically, \(r_{\rm relax}(10\Gyr)\) is about 6-10 times our adopted
softening length, increasing with particle number. As such, at full
resolution, we can only trust density profiles down to \(\sim10\) times
our softening length, sufficient to resolve stellar density profiles.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/iso_converg_num.png}}
\caption[Numerical halo convergence]{Numerical convergence test for
circular velocity as a function of log radius for simulations with
different total numbers of particles in isolation. Residuals in bottom
panel are relative to NFW. The initial conditions are dotted and the
converged radius is marked by arrows (Eq.~\ref{eq:t_relax}). Note that a
slight reduction in density starting around \$r = 30 \$kpc is expected
given our truncation choice.}\label{fig:numerical_convergance}
\end{figure}

@@

To perform the simulations of a given galaxy in a given potential, we
centre the isolation run's final snapshot
(Section~\ref{sec:shrinking_spheres}) and place the dwarf galaxy in the
specified orbit in the given potential. We typically run the simulation
for 10 Gyr, which allows us to orbit slightly past the expected initial
conditions.

\section{Post processing}\label{post-processing}

\subsection{Centring}\label{sec:shrinking_spheres}

@@

\begin{itemize}
\tightlist
\item
  Circular velocity is computed assuming spherical symmetry and only
  shown for every 200th particle (ranked from the centre outwards)
\item
  \(v_{\rm circ\ max}\) and \(r_{\rm circ,\ max}\) is found by
  least-squares fitting the NFW functional velocity form to the points
  of the velocity profile that have the 10\% highest velocities. This is
  not a necessarily good fit, especially as the halo becomes stripped,
  but accurate enough to find a reliable maximum.
\end{itemize}

\subsection{Stellar probabilities}\label{stellar-probabilities}

We ``paint'' stars onto dark matter particles using the particle tagging
method \citep[e.g.][]{bullock+johnston2005}, assuming spherical
symmetry. Let \(\Psi\) be the potential (normalized to vanish at
infinity) and \({\cal E}\) is the binding energy
\({\cal E} = \Psi - 1/2 v^2\). If we know \(f({\cal E})\), the
distribution function (phase-space density in energy), then we assign
the stellar weight for a given particle with energy \({\cal E}\) is

\begin{equation}{
P_\star({\cal E}) = \frac{f_\star({\cal E})}{f_{\rm halo}({\cal E})}.
}\end{equation} While \(f({\cal E})\) is a phase-space density, the
differential energy distribution includes an additional \(g({\cal E})\)
occupation term (BT87). We use Eddington inversion to find the
distribution function, (eq. 4-140b in BT87)

\begin{equation}{
f({\cal E}) = \frac{1}{\sqrt{8}\, \pi^2}\left( \int_0^{\cal E} \frac{d^2\rho}{d\Psi^2} \frac{1}{\sqrt{{\cal E} - \Psi}}\ d\Psi + \frac{1}{\sqrt{\cal E}} \left(\frac{d\rho}{d\Psi}\right)_{\Psi=0} \right).
}\end{equation}

In practice the right, boundary term is zero as \(\Psi \to 0\) as
\(r\to\infty\), and if \(\rho \propto r^{-n}\) at large \(r\) and
\(\Psi \sim r^{-1}\) then \(d\rho / d\Psi \sim r^{-n+1}\) which goes to
zero provided that \(n > 1\). We take \(\Psi\) from the underlying
assumed analytic dark matter potential. \(\rho_\star\) is pre-specified
or calculated via Abel integration / deprojection.

@@

Perhaps the simplest in form, the 2D exponential is commonly used to
describe dwarf galaxy's density profiles as well as stellar disks. The
stellar profile is given by \begin{equation}{
\Sigma_\star(R) = A e^{-R / R_s}
}\end{equation} where \(A\) is the normalization and \(R_s\) is the
exponential scale radius. \#\#\# Plummer

The Plummer density profile is generally considered to be a reasonable
empirical fit to Globular Clusters, where it was first proposed
\citep{sadfjk}. Plummer profiles are also sometimes used to describe
dwarf spheroidal galaxies REFS and galactic bulges REF. A Plummer
profile is defined by \begin{equation}{
\Sigma(R) = \frac{M}{4\pi R_s^2} \left(1 + \frac{R^2}{R_s^2}\right)^{-2} ,
}\end{equation} where \(M\) is the total mass and \(R_s\) is the
characteristic scale radius.

@@

Stellar velocity dispersion is calculated for all stars within a 1kpc
sphere of the centre. We assume
\(\sigma_{\rm los} = \sqrt{\sigma_x^2 + \sigma_y^2 + \sigma_z^2}/ \sqrt 3\),
i.e.~isotropic velocity dispersion.

We calculate density profiles similar to stars and assume a constant bin
width in \(\log R\) of \(2 {\rm IQR} / \sqrt[3]{N}\) (Freedman-Diaconis
prescription). \(R\) may be either the cylendrical radius in \(x-y\) or
the on-sky project \(\xi, \eta\) tangent plane coordinates.

Because our dwarfs are assumed to be spherical/isotropic, we retain this
assumption when calculating predicted density profiles.

### 2025-07-15 thesis/numerical_convergence.tex
major thesis restructure

@@

\subsubsection{Softening}\label{softening}

\citet{power+2003} suggest the empirical rule that the ideal softening
(balancing integration time and only compromising resolution in
collisional regime) is \[
h_{grav} = 4 \frac{R_{200}}{\sqrt{N_{200}}}
\]

For our isolation halo (\(M_s=2.7\), \(r_s=2.76\)) and with \(10^7\)
particles, this works out to be \(0.044\,{\rm kpc}\).We adpoted the
slightly smaller softening which was reduced by a factor of
\(\sqrt{10}\) which appears to improve agreement slightly in the
innermost regions. However, this choice likely unnecessarily increases
computation time for the relative gain in accuracy.

@@

\subsubsection{Timestepping and force
accuracy}\label{timestepping-and-force-accuracy}

@@

\subsubsection{Alternative methods}\label{alternative-methods}

\begin{itemize}
\tightlist
\item
  FMM
\item
  PMM-tree
\item
  Gadget2
\item
  etc.
\end{itemize}

@@

\subsubsection{Fiducial Parameters}\label{fiducial-parameters}

>>>>

\chapter{Numerical convergence and
parameters}\label{sec:extra_convergence}

@@

\section{Softening}\label{softening}

One challenge of N-body integration is close, collisional encounters,
assuming point particles, cause divergences in the local force. However,
this should not occur in a \emph{collisionless} simulation. As a result,
most collisionless N-body codes adopt a gravitational softening, a
length scale below which the force of gravity begins to decrease between
point particles.

\citet{power+2003} empirically suggest that the ideal softening is
\begin{equation}{
h_{\rm grav} = 4 \frac{R_{200}}{\sqrt{N_{200}}},
}\end{equation}

where \(h_{\rm grav}\) is the softening length, and \(N_{200}\) is the
number of particles within \(R_{200}\). This choice balances integration
time and only compromises resolution in collisional regime.

For our isolation halo (\(M_s=2.7\), \(r_s=2.76\)) and with \(10^7\)
particles, this works out to be \(0.044\,{\rm kpc}\).We adpoted the
slightly smaller softening which was reduced by a factor of
\(\sqrt{10}\) which appears to improve agreement slightly in the
innermost regions.

@@

\section{Time stepping and force
accuracy}\label{time-stepping-and-force-accuracy}

@@

@@

\section{Fiducial parameters}\label{fiducial-parameters}

### 2025-07-15 thesis/sculptor.tex
major thesis restructure

\section{Sculptor (MW only)}\label{sculptor-mw-only}

\subsection{Orbital estimation}\label{orbital-estimation}

fig.~\ref{fig:unref} contains example orbits

@@

\subsection{Simulations}\label{simulations}

@@

\section{Sculptor (LMC orbits)}\label{sculptor-lmc-orbits}

As discussed in \citet{battaglia+2022}, Sculptor's orbit is strongly
influenced by the presence of an LMC. Figure
fig.~\ref{fig:scl_lmc_orbit_effect}, the addition of an LMC reduces
Scl's pericentre with the MW and implies that Scl may be on its first
infall with the MW, in contrast to the discussion above. Thus, the LMC
has a critical impact on the evolution of Scl. Additionally, Scl passes
rapidly but relatively close to the LMC in the past 100 Myr.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250415112706417.png}}
\caption{image-20250415112706417}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250415113122385.png}}
\caption{image-20250415113122385}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250415113133353.png}}
\caption{image-20250415113133353}
\end{figure}

>>>>

\section{Sculptor}\label{sculptor}

\subsection{Derived properties}\label{derived-properties}

\begin{table*}[t]
\centering
\begin{tabular}{lll}
\toprule
parameter & value & reference\\
\midrule
$L_\star$ & $1.82_{-0.22}^{+0.25}\times10^6\ L_\odot$ & \\
$M_\star$ & $2.7_{-0.6}^{+0.7} \times10^6\ {\rm M}_\odot$ & \\
$M_{200}$ & $0.48_{-0.25}^{+0.52}\ M_0$ & \\
$c_{\rm NFW}$ & $13.1_{-2.8}^{+3.6}$ & \\
$v_{\rm circ, max}$ & $31\pm8\,\kms$ & \\
$r_{\rm circ, max}$ & $5.9 \pm 2.9$ kpc & \\
$r_{\rm break, obs}$ & $25 \pm 5$ arcmin & \\
$t_{\rm break, obs}$ & $110\pm30$ Myr & \\
\bottomrule
\end{tabular}
\end{table*}

\subsection{Milky Way tides}\label{milky-way-tides}

\subsubsection{Orbital properties}\label{orbital-properties}

Fig.~\ref{fig:unref} contains example orbits

@@

\subsubsection{Tidal effects}\label{tidal-effects}

@@

\subsection{Effects of the LMC}\label{effects-of-the-lmc}

\subsubsection{Orbital effects}\label{orbital-effects}

As discussed in \citet{battaglia+2022}, Sculptor's orbit is strongly
influenced by the presence of an LMC. Figure
Fig.~\ref{fig:scl_lmc_orbit_effect}, the addition of an LMC reduces
Scl's pericentre with the MW and implies that Scl may be on its first
infall with the MW, in contrast to the discussion above. Thus, the LMC
has a critical impact on the evolution of Scl. Additionally, Scl passes
rapidly but relatively close to the LMC in the past 100 Myr.

Finally, we also consider the influence of the large Milky Cloud on the
orbits and evolution of Scl. We adopt the \citet{vasiliev2024} multipole
approximation of an N-body simulation of the LMC and MW. Their initial
conditions are

\begin{itemize}
\tightlist
\item
  MW halo:
\item
  MW bulge (static):
\item
  MW disk (static):
\item
  LMC halo:
\end{itemize}

@@

\begin{figure}
\centering
\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250415112706417.png}
\caption[Effect of LMC on Sculptor's Orbit]{Orbits of Sculptor with and
without the LMC. The potential is that of \citet{vasiliev+2021} in each
case.}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250415113122385.png}
\caption[Selected orbits of Sculptor with an LMC]{Selected orbits of
Sculptor in the MW frame. \textbf{TODO}: include LMC distance or frame,
maybe xyz instead?}
\end{figure}

\subsubsection{Tidal effects}\label{tidal-effects-1}

\section{Ursa Minor}\label{ursa-minor}

\subsection{Derived properties}\label{derived-properties-1}

\begin{table*}[t]
\centering
\begin{tabular}{lll}
\toprule
parameter & value & reference\\
\midrule
$L_\star$ &  & \\
$M_\star$ &  & \\
$M_{200}$ &  & \\
$c_{\rm NFW}$ &  & \\
$v_{\rm circ, max}$ & $\,\kms$ & \\
$r_{\rm circ, max}$ & kpc & \\
$r_{\rm break, obs}$ & $30 \pm 5$ arcmin & \\
$t_{\rm break, obs}$ & $120\pm30$ Myr & \\
\bottomrule
\end{tabular}
\end{table*}

\subsection{Orbital properties}\label{orbital-properties-1}

\begin{table*}[t]
\centering
\begin{tabular}{lllllllll}
\toprule
Orbit & Mean &  & smallperi &  &  &  &  & \\
\midrule
ra &  &  &  &  &  &  &  & \\
dec &  &  &  &  &  &  &  & \\
distance &  &  &  &  &  &  &  & \\
pmra &  &  &  &  &  &  &  & \\
pmdec &  &  &  &  &  &  &  & \\
rv &  &  &  &  &  &  &  & \\
x &  &  &  &  &  &  &  & \\
y &  &  &  &  &  &  &  & \\
z &  &  &  &  &  &  &  & \\
vx &  &  &  &  &  &  &  & \\
vy &  &  &  &  &  &  &  & \\
vz &  &  &  &  &  &  &  & \\
\bottomrule
\end{tabular}
\end{table*}

\subsection{Tidal effects}\label{tidal-effects-2}
